---
title: "Intro to Programming Concepts"
author: "Nikki Rubinstein"
date: "Saturday, June 25, 2016"
output: ioslides_presentation
---

```{r, include=FALSE}
library(jpeg)
library(png)
library(grid)
```

## Telling your computer what to do
<!-- concept of programming languages -->
<!-- play simon says -->
```{r fig.width=5, fig.height=5,echo=FALSE, fig.align='center'}
img <- readJPEG("fig//Olate_dogs_-_on_wheel.jpeg")
grid.raster(img) 
``` 

<!-- image source: https://upload.wikimedia.org/wikipedia/commons/d/d8/Olate_dogs_-_on_wheel.jpeg -->

## Interacting with your computer
<!-- introduce the idea of variables, assignment and memory storage
at this point people should name their teams / tables -->
```{r fig.width=5, fig.height=5,echo=FALSE, fig.align='center'}
img <- readPNG("fig//Abstract_art_(PSF).png")
grid.raster(img) 
``` 
<!-- image source: https://upload.wikimedia.org/wikipedia/commons/6/66/Abstract_art_(PSF).png -->

## Interacting with your computer

```{r fig.width=5, fig.height=5,echo=FALSE, fig.align='center'}
img <- readJPEG("fig//pexels-photo.jpg")
grid.raster(img) 
``` 
<!-- image source: https://static.pexels.com/photos/15393/pexels-photo.jpg -->

## What types of information does your computer understand?
<!-- These are the building blocks of our conversation with the computer, much like words are the building blocks of a book -->
* **numeric** 
    + integers, e.g. `1, 2, 3`
    + doubles, e.g. `1.25, 3.78`
    + complex numbers, e.g. `3 + 2i`
* **character or string** 
    + e.g. `"Research", "Platforms"`
* **logical** 
    + e.g. `TRUE` and `FALSE`

## How can we group information?



<!-- much like how we create sentences from words -->
* vectors / tuples
* matrices / data.frames / arrays
* lists 
* dictionaries

## vectors / tuples
```{r fig.width=5, fig.height=5,echo=FALSE, fig.align='center'}
img <- readJPEG("fig//lechner-50119_960_720.jpg")
grid.raster(img) 
``` 
<!-- image source: https://pixabay.com/static/uploads/photo/2012/06/16/20/52/lechner-50119_960_720.jpg -->

## matrices / data.frames / arrays
```{r fig.width=5, fig.height=5,echo=FALSE, fig.align='center'}
img <- readJPEG("fig//House_of_Air_The_Matrix.jpg")
grid.raster(img) 
``` 
<!-- image source: https://upload.wikimedia.org/wikipedia/commons/f/fa/House_of_Air_The_Matrix.jpg -->

## lists
```{r fig.width=5, fig.height=5,echo=FALSE, fig.align='center'}
img <- readJPEG("fig//musical instrument collection.jpg")
grid.raster(img) 
``` 
<!-- image source: http://s0.geograph.org.uk/geophotos/03/54/79/3547925_df1f2e83.jpg -->

## dictionaries
```{r fig.width=5, fig.height=5,echo=FALSE, fig.align='center'}
img <- readJPEG("fig//dictionary.jpg")
grid.raster(img) 
``` 
<!-- image source: https://pixabay.com/static/uploads/photo/2015/12/23/22/16/collaboration-1106196_960_720.jpg -->

## How to add flexibility to the conversation
**If statements**
```{r fig.width=5, fig.height=5,echo=FALSE, fig.align='center'}
img <- readJPEG("fig//Am I hungry.jpg")
grid.raster(img) 
``` 

## How to add flexibility to the conversation
**For loops**
```{r fig.width=5, fig.height=5,echo=FALSE, fig.align='center'}
img <- readJPEG("fig//Man_and_woman_dancing_a_waltz.jpg")
grid.raster(img) 
``` 
<!-- image source: https://upload.wikimedia.org/wikipedia/commons/4/49/Man_and_woman_dancing_a_waltz_(rbm-QP301M8-1887-197a~6).jpg -->

## How to add flexibility to the conversation
**Functions**
```{r fig.width=5, fig.height=5,echo=FALSE, fig.align='center'}
img <- readJPEG("fig//cooking.jpg")
grid.raster(img) 
``` 
<!-- image source: https://pixabay.com/static/uploads/photo/2015/10/30/05/48/cooking-1013455_960_720.jpg -->
